---
title: "Describing a numeric variable grouped by categories"
output:
  html_document:
    css: tutorial.css
    fig_caption: yes
    highlight: pygments
    theme: simplex
    toc: yes
    toc_float: yes
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = T)
library(tigerstats)
lattice.options(default.theme = standard.theme(color = FALSE))
```
This page was last updated on `r format(Sys.time(), '%B %d, %Y')`.

* * *

## Getting started  

### Import the data

Import the `students.csv` data set:
```{r read_data}
students <- read.csv(url("https://people.ok.ubc.ca/jpither/datasets/students.csv"), header = TRUE)
```

* * *

### Load the required packages

Load the required `tigerstats` package:
```{r attach_libraries}
library(tigerstats)
```

* * *

## Describing a numerical variable grouped by categories

When we are visualizing how a numeric response variable is associated with a categorical explanatory variable, it is advisable to calculate descriptive statistics for the response variable grouped by category.  For instance, what is the mean and standard deviation of height for each of the sexes?
 
We can do this using the `favstats` function:
```{r summary_stats_group}
favstats(height_cm ~ sex, data = students)
```

__TIP__: Note the syntax, with the response variable on the left of the `~` symbol, and the the explanatory (categorical) variable on the right.

The `favstats` function provides all of the descriptive statistics you'd typically want to report, and more than you need. However, you can use the other functions we learned in a similar way:
```{r desc_stats_group}
mean(height_cm ~ sex, data = students)
```

* * *

## List of functions (and the source packages) used in tutorial

__Getting started__:

* `read.csv`
* `url`
* `library`  

__Descriptive stats__:

* `favtstats`
* `mean`



