---
title: "The normal distribution"
output:
  html_document:
    css: tutorial.css
    fig_caption: yes
    highlight: pygments
    theme: simplex
    toc: yes
    toc_float: yes
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
library(tigerstats, warn.conflicts = FALSE, quietly = TRUE)
lattice.options(default.theme = standard.theme(color = FALSE))
```
This page was last updated on `r format(Sys.time(), '%B %d, %Y')`.

* * *

## Getting started

In this tutorial we will learn about the following:

* constructing a Gaussian (normal) distribution
* the **Central Limit Theorem**
* the **standard normal distribution**
* calculating *Z*-scores
* calculating *P*-values using a normal distribution

* * *

### Required packages

* `tigerstats`

```
library(tigerstats)
```

* * *

### Required data

* the "flu" dataset.  These are the data associated with the Spanish Flu example in Chapter 10 of the text.  

```{r}
flu <- read.csv(url("https://people.ok.ubc.ca/jpither/datasets/flu.csv"), header = TRUE)
inspect(flu)
head(flu)
```


* * *

## The Gaussian (normal) distribution

The formula for a normal or "Gaussian" distribution is:  

```{r fig.width = 2, fig.height = 0.5, echo = FALSE}
library(png)
img <- readPNG("./more/normal_equation.png")
grid.raster(img)
```

To characterize a normal distribution, one needs only two parameters: $\mu$ and $\sigma$.  

* * *

## Generating a normal distribution

The `rnorm` function provides the means to generate `n` values drawn from a population with a specific mean and standard deviation.  
```
?rnorm
```

Using this function we can essentially pretend that, in the background, R creates a vast population in which the variable of interest exhibits a normal distribution with the parameter values ($\mu$ and $\sigma$) we specify. Then, we take a random sample of size _n_ from this population.  

This function, as with similar functions you've used previously, includes an algorithm that relies on a random number generator. So, if we want to ensure that we all get the same set of values in our sample, we can use the `set.seed` command, which sets the seed value for the program's random number generator.  

```{r}
set.seed(50)
example.norm <- rnorm(n = 100, mean = 5, sd = 1.5)
```

Let's also visualize the resulting frequency distribution with a histogram:

```{r fig.cap = 'Figure 1: Histogram of n = 100 values drawn from a normal distribution with mean = 5 and sd = 1.5', fig.width = 5, fig.height = 4}
histogram(~ example.norm, 
          type = "count", 
          col = "grey",
          xlab = "Value",
          ylab = "Frequency",
          main = "")
```

* * *

1. Generating normal distributions

Repeat the above commands using a few different sample sizes (*n*) and values for the mean and standard deviation.  

Knowing how to generate data with known characteristics (e.g. normal distribution with particular mean and sd), is a handy skill.  Often we wish to test out ideas with such "fake" or "made-up" data, specifically data that exhibit particular properties.  

* * *

## The Central Limit Theorem - a simulation

For this example we'll use the flu dataset described in example 10.6 in the text (page 286).  

Let's look at a histogram showing the frequency distribution of the ages at death in Switzerland in 1918, and make it look like the figure in the text book:  

```{r fig.cap = 'Figure 2: Frequency distribution of age at death for 75034 people in Switzerland in 1918 during the Spanish flu epidemic', fig.height = 4, fig.width = 5}
histogram(~ age, data = flu, 
          breaks = seq(from = 0, to = 102, by = 2), # this provides the breaks for the bins
          type = "count", 
          col = "firebrick",
          xlab = "Age at death (yrs)",
          ylab = "Frequency",
          main = "")
```
  

You can see that there are multiple modes in this frequency distribution, and overall the distribution is highly non-normal.  

According to the **Central Limit Theorem**, the sum or mean of a large number of measurements randomly sampled from a non-normal population is approximately normally distributed.  

Here we will build on what we learned in the [Sampling, Estimation, and Uncertainty](https://people.ok.ubc.ca/jpither/modules/Sampling_Estimation_Uncertainty.html) tutorial to help gain an appreciation of the Central Limit Theorem. 

The code below first sets the seed, then defines a sample size (_n_) to use when drawing a random sample from the flu dataset.  

It then tells R the **number of trials** for our simulation, 10000.

The `sample` command is taking a random sample (without replacement) of size _n_ from the ages in the dataset.  

The `mean` command is calculating the mean of the ages within the given sample.  

The `do` function is telling R to repeat the `mean` and `sample` procedure **number of trials** times, and to put the resulting set of means in a vector called `samp.data`.  

The procedure will take 10 or so seconds to complete, so be patient.  

```{r}
set.seed(42) # set the seed
n <- 20  
number.of.trials <- 10000
samp.data <- do(number.of.trials) * mean(sample(flu$age, size = n, replace = FALSE))
```

In the [Sampling, Estimation, and Uncertainty](https://people.ok.ubc.ca/jpither/modules/Sampling_Estimation_Uncertainty.html) tutorial, we used the `sapply` function to repeat the sampling procedure 10000 times... here we use the much handier `do` function.  

Now produce a histogram showing the freqency distribution of the 10000 sample means.  

Be sure to include defined x- and y-axis limits (the `xlim` and `ylim` arguments), so that subsequent histograms are shown on the same scale.

```{r fig.cap = 'Figure 3: Frequency distribution of 10000 sample means for the variable age at death.', fig.width = 4, fig.height = 4.5}
histogram(~ samp.data, 
          type = "count",
          xlim = c(-10, 100),
          ylim = c(0, 3000),
          col = "firebrick",
          las = 1,
          xlab = "Mean age (yrs)",
          ylab = "Frequency",
          main = paste("Sample size = ", n, sep =" ")) # show "n" in heading
```


* * *

1. Simulate central limit theorem

* Repeat the above commands, but change _n_ to 2.  
* Repeat the commands again, but change _n_ to 40.  

What happens to the frequency distribution in each case?  
Compare your frequency distribution of sample means to the histogram showing all the age data (Figure 2).  

* * *

## The standard normal distribution 

The Gaussian (normal) distribution is a **continuous probability distribution**.  

The **standard normal distribution** is a distribution that has a mean $\mu$ = 0 and standard deviation $\sigma$ = 1.  

Any variable that exhibits a normal distribution in the population can be re-scaled to fit a standard normal distribution, so long as we know the population mean ($\mu$) and standard deviation ($\sigma$).  

Why would we wish to re-scale a variable to fit a standard normal distribution?  

Doing so allows us to:  

* easily quantify how far a particular value is from the population mean, in units of "standard deviation" - a useful common currency for describing the unusualness of values.  We call this quantity a **Z-score**, or a **standard normal deviate** 
* calculate a probability value, i.e. *P*-value, associated with a particular **Z-score** 

* * *

### Calculating *Z*-scores

Consider the following example: 

The height of adult males in Great Britain is normally distributed, and has $\mu$ = 177.0 cm and $\sigma$ = 7.1 cm.  

My uncle is 192cm tall. He is clearly tall compared to the average height of males in the population. But how much taller?  

We can calculate how far his height is from the population mean in units of standard deviations.  

To do so by hand, we use this formula:  

$$Z = \frac{Y - \mu}{\sigma}$$

So, plugging in the numbers:  

$$Z = \frac{192 - 177}{7.1}$$ 
$$Z = 2.11$$   

So my uncle's height is 2.11 standard devations above the mean height for British males.  


* * *

### Calculating probabilities with a normal distribution

Once we have calculated a *Z*-score, we can calculate the *P*-value associated with that *Z*-score.  

Let's re-visit the male height example from above.  

MI-5 (the spy agency of the British government) does not want male spies who are taller than 180.3 cm (they would stick out in a crowd too much).  

What proportion of British adult men are excluded from a career as a spy for MI-5?  

That is, what is Pr[_height_ > 180.3] ?  

In R, we can use the `pnormGC` function from the `tigerstats` package (a wrapper of the `pnorm` function in the base package) to calculate probabilities from a normal distribution, and visualize the area under the curve.  

```
library(tigerstats)
?pnormGC
```

You can find out more about the `pnormGC` function at the TigerStats website [here](http://homerhanumat.com/tigerstats/pnormGC.html).  

So, for the current example:  

```{r fig.width = 5, fig.height = 4}
pnormGC(180.3, region = "above", mean = 177, sd = 7.1, graph = TRUE)
```

We see that about `r round(100*pnormGC(180.3, region = "above", mean = 177, sd = 7.1, graph = FALSE),0)` percent of British males would be excluded from service as a MI-5 spy.  

R does the conversion to a *Z*-score for us, and calculates the area under the curve.  

If we didn't have a computer, we could look up areas under the curve from a **Table of Z-scores**, such as the one in the text on page 706-707.  

* * *

### Calculating probabilities of sample means

The standard normal distribution can be used to calculate the probability of drawing a sample with a mean in a given range, assuming we know $\mu$ and $\sigma$:

$$Z = \frac{\bar{Y} - \mu}{\sigma_\bar{Y}}$$

The quantity $\sigma_\bar{Y}$ is the standard deviation of the sampling distribution of $\bar{Y}$, also known as the **standard error of the mean**.  

**Example**:  

Baby weights are normally distributed in the population and have a mean $\mu$ = 3339g and standard deviation $\sigma$ = 573g.  

What is the probability that a random sample of *n* = 80 babies would yield a mean weight of at least 3370g?  

In other words, what is Pr[$\bar{Y}$ > 3370]?  

Here, $$\sigma_\bar{Y} = \frac{\sigma}{\sqrt{n}} = \frac{573}{\sqrt{80}} = 64.1$$  

From this we calculate the *Z*-score:

$$Z = \frac{\bar{Y} - \mu}{\sigma_\bar{Y}} = \frac{3370 - 3339}{64.1} = 0.48$$

In other words, Pr[$\bar{Y}$ > 3370] = Pr[*Z* > 0.48]  

We can look up the associated probability value in Table B in the text (p 706).  

Alternatively, we can calculate the probability in R using the `pnormGC` function, and making sure to enter the **standard error of the mean** in the `sd` argument, rather than the standard deviation (because here we're dealing with the distribution of means):  

```{r fig.width = 5, fig.height = 4}
pnormGC(3370, region = "above", mean = 3339, sd = 64.1, graph = TRUE)
```

Thus, Pr[$\bar{Y}$ > 3370] = `r round(pnormGC(3370, region = "above", mean = 3339, sd = 64.1, graph = FALSE), 2)`.  


* * *

1. Calculate a probability  

MI-5 also discriminates against shorter people... no spies can be less than 173.7cm tall.  

* What proportion of the male population in Great Britain qualify, with respect to height, for being a spy?  

That is, what is Pr[173.7 > _height_ > 180.3] ?  

**HINT**: look at the available options for the `region` argument to the `pnormGC` function 

* * *

### Calculating percentiles from a normal population

Consider the following question: What height is associated with the first quartile in the population?  

In notation format, this looks like:  

Pr[_height_ < X] = 0.25

For this problem, we are given the probability *P*, and need to find the corresponding value of *Z*.  

For this, we use the `qnormGC` function from the `tigerstats` package (a wrapper of the `qnorm` function in the base package):  

```
?qnormGC
```

```{r fig.width=5, fig.height = 4}
qnormGC(area = 0.25, region = "below", mean = 177, sd = 7.1, graph = TRUE)
```

We see that `r round(qnormGC(area = 0.25, region = "below", mean = 177, sd = 7.1, graph = TRUE),1)`cm is the height associated with the first quartile in the population.  

* * *

## List of functions

__Getting started__:

* `read.csv`
* `url`
* `library` 

__Data frame structure__: 

* `head`
* `inspect` (`tigerstats` / `mosaic` packages)

__Simulation__:

* `set.seed`
* `sample`
* `do`
* `mean`

__Normal distribution__:

* `rnorm`
* `pnormGC` (`tigerstats` package)
* `qnormGC` (`tigerstats` package)

__Graphs__:

* `histogram`
