---
title: "Least-squares linear regression"
output:
  html_document:
    css: tutorial.css
    fig_caption: yes
    highlight: pygments
    theme: simplex
    toc: yes
    toc_float: yes
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(eval = TRUE, echo = TRUE, cache = TRUE)
```
This page was last updated on `r format(Sys.time(), '%B %d, %Y')`.

* * *

## Getting started

In this tutorial you will:  

* learn about using least-squares linear regression to model the relationship between two numeric variables, and to make predictions

* * *

### Required packages

* `tigerstats`
* `visreg`  

The latter package is likely new to you, so install it first by typing the following in the command console:  

```
install.packages("visreg")
```

Load the packages:  

```{r, message = FALSE, warning = FALSE}
library(tigerstats)
library(visreg)
```

* * *

### Required data

The "plantbiomass" dataset:  

```{r}
plantbiomass <- read.csv(url("https://people.ok.ubc.ca/jpither/datasets/plantbiomass.csv"), header = TRUE)
```

* * *

### Data exploration

The `plantbiomass` (see p. 551 in text) dataset includes data describing plant biomass measured after 10 years within each of 5 experimental "species richness" treatments, wherein plants were grown in groups of 1, 2, 4, 8, or 16 species. The treatment variable is called `nSpecies`, and the response variable `biomassStability` is a measure of ecosystem stability. The research hypothesis was that increasing diversity (species richness) would lead to increased ecosystem stability.  

```{r}
head(plantbiomass)
inspect(plantbiomass)
```

We see that both variables are numeric, and that there are 161 observations overall.  

* * *

## Regression analysis

When we are only interested in the strength of a linear association between two numerical variables, we use a correlation analysis.  

When we are interest in making _predictions_, we use regression analysis.  

Often in the literature you see authors reporting regression results when in fact there was no rationale for using regression; a correlation analysis would have been more appropriate.  

The two analyses are mathematically related.  

Regression analysis is extremely common in biology, and unfortunately it is also common to see incorrect implementation of regression analysis. In particular, one often sees scatterplots that clearly show violations of the assumptions of regression analysis (see below).  **Failure to appropriately check assumptions can lead to misleading and incorrect conclusions**.   

* * *

### Hypothesis testing - not really!

Recall from your high-school training that the equation for a straight line is:  

_Y_ = _a_ + _bX_  

In regression analysis, the "least-squares line" is the line for which the sum of all the _squared deviations in Y_ is smallest.  

In a regression context, _a_ is the Y-intercept and _b_ is the slope of the regression line.  

Regression analysis falls under the heading of **inferential statistics**, which means we use it to draw inferences about a linear relationship between Y and X within a _population_ of interest. So in the actual population, the true least-squares line is represented as follows:  

_Y_ = $\alpha$ + $\beta$_X_  

In practice, we estimate the "parameters" $\alpha$ and $\beta$ using a random sample from the population, and by calculating _a_ and _b_ using regression analysis.  

**NOTE**: the $\alpha$ in the regression equation has no relation to the $\alpha$ that sets the significance level for a test!  

In least-squares regression analysis, the null hypothesis is that the _slope_ of the relationship between Y and X is equal to zero. That is:  

**H~0~**: $\beta$ = 0  
**H~A~**: $\beta$ $\neq$ 0     

However, one does not typically state the null and alternative hypotheses when conducting a regression analysis. Rather, they are **implied**.  

As such, **there is no need to explicitly state these hypotheses for a regression analysis**.  

You must, however, know what is being tested: the null hypothesis that the slope is equal to zero.  

* * *

### Steps to conducting regression analysis

In regression analysis one must conduct the actual analysis before one can check the assumptions. Thus, the order of operations is a bit different from that of other tests.  

**Follow these steps when conducting regression analysis:**

* Set an $\alpha$ level (default is 0.05)  
* Provide an appropriate figure, including figure caption, to visualize the data  
* Provide a line or two interpreting your figure, and this may inform your concluding statement  
* Conduct the regression analysis, save the output, and use the **residuals** from the analysis to check assumptions  
* Assumptions  
    + state the assumptions of the test  
    + use appropriate figures and / or tests to check whether the assumptions of the statistical test are met
    + transform data to meet assumptions if required
    + if assumptions can't be met (e.g. after transformation), use alternative methods (**you are not required to know these alternative methods**)
    + if data transformation is required, then do the transformation and provide another appropriate figure, including figure caption, to visualize the transformed data, and a line or two interpreting this new figure  
* Calculate the confidence interval for the slope and the *R*^2^ value, report these in concluding statement  
* if the regression is significant, provide a scatterplot that includes the regression line and confidence bands, with appropriate figure caption  
* Use the output from the regression analysis (above) draw an appropriate conclusion and communicate it clearly  

* * *

### Plant biomass example

Using the plant biomass data, let's test the null hypothesis that ecosystem stability cannot be predicted from plant species richness.  

* * *

### Visualize the data      

We learned in an earlier [tutorial](https://people.ok.ubc.ca/jpither/modules/Visualizing_association_two_variables.html#visualizing_association_between_two_numeric_variables) that the best way to visualize an association between two numeric variables is with a scatterplot, and that we can create a scatterplot using the `plot` function:  

```{r fig.cap = "Figure 1: Stability of biomass production over 10 years in 161 plots and the initial number of plant species assigned to plots.", fig.width = 4.2, fig.height = 4}
plot(biomassStability ~ nSpecies, data = plantbiomass, 
           xlab = "Species number treatment",
           ylab = "Biomass stability",
           pch = 1,  # pch changes the symbol type
           col = "firebrick",
           las = 1)  # orients y-axis tick labels properly
```

**NOTE**: This figure that you first produce for visualizing the data may or may not be necessary to present with your results. If the regression analysis ends up being significant, then you'll create a new figure to reference in your concluding statement (see below). If your regression analysis is non-significant, then you should present and refer to this figure (which does not include a best-fit regression line).  

In Figure 1 we do not report units for the response variable, because it does not have units

* * *

### Interpreting a scatterplot

In an earlier [tutorial](https://people.ok.ubc.ca/jpither/modules/Visualizing_association_two_variables.html#interpreting_and_describing_a_scatterplot), we learned how to properly interpret a scatterplot, and **what information should to include in your interpretation**. Be sure to consult that tutorial.   

> We see in Figure 1 that there is a weak, positive, and somewhat linear association between biomass stability and species number treatment.  There appears to be increasing spread of the response data with increasing values of the independent (x) variable, and perhaps outliers in the highest group of species richness. We should keep this in mind.   

* * *

### Assumptions of regression analysis

Regression analysis assumes that:  

* the true relationship between X and Y is linear  
* for every value of X the corresponding values of Y are normally distributed  
* the variance of Y-values is the same at all values of X
* at each value of X, the Y measurements represent a random sample from the population of possible Y values  

* * *

### Checking the assumptions of regression analysis

Strangely, we must conduct the regression analysis before checking the assumptions. This is because we check the assumptions using the **residuals** from the regression analysis.  

We therefore implement the regression analysis and be sure to store the output for future use.   

We'll use an $\alpha$ level 0.05. 

The function to use is the `lm` function, which we saw in [tutorial 15](https://people.ok.ubc.ca/jpither/modules/Comparing_more_than_2_means.html#conduct_the_anova_test), but here our independent (explanatory) variable is a numeric variable rather than a categorical variable.  

Let's run the regression analysis and assign the output to an object as follows:  

```{r}
biomass.lm <- lm(biomassStability ~ nSpecies, data = plantbiomass)
```

Before we do anything else (e.g. look at the results of the regression), we need to first check the assumptions!  

**TIP**: The `plot.lm` function in the base stats package can be used on the regression output object to produce a series of so-called "regression diagnostic plots".  See the help file for info.  

To get the residuals, which are required to check the assumptions, you simply use the `residuals` function (from the base stats package) on the object holding the regression output from above (`biomass.lm`). We'll create a new variable (`biomass.lm.resids`) in the dataframe `plantbiomass` to store these residuals:  

```{r}
plantbiomass$biomass.lm.resids <- residuals(biomass.lm)
```

"Residuals" represent the vertical difference between each observed value of Y (here, biomass stability) the least-squares line (the predicted value of Y) at each value of X.  

You might think that one could check the normality assumption by simply conducting usual diagnostics on the raw Y values. However, the appropriate method is to use the **residuals** from the regression, and check that these are normally distributed.  

We can check the normality of residuals assumption with a normal quantile plot, which you've seen [previously](https://people.ok.ubc.ca/jpither/modules/Comparing_two_means.html#check_assumptions16).  


```{r fig.cap = "Figure 2: Normal quantile plot of the residuals from a regression of biomass stability on species number for 161 plots", fig.width = 4.2, fig.height=4}
{
qqnorm(plantbiomass$biomass.lm.resids, las = 1)
qqline(plantbiomass$biomass.lm.resids)
}
```

**TIP**: The outer curly brackets in the code above ensure that the `qqline` function is implemented properly and added to the main quantile plot in the R chunk (it avoids an error).  

> Figure 2 shows that the residuals don't really fall consistently near the line in the normal quantile plot; there is curviture, and increasing values tend to fall further from the line. This suggests the need to **log-transform the data**.  

Before trying a transformation, let's also check the assumption that (i) the variance of Y is the same at all values of X, and (ii) the assumption that the association is linear.  To do this, we plot the residuals against the original "X" variable, which here is the number of species initially used in the treatment. Note that we'll add a horizontal line ("segement") at zero for reference:  

```{r fig.cap = "Figure 3: Residual plot from a regression of biomass stability on species number for 161 plots.", fig.width = 4.2, fig.height = 4}
{
plot(biomass.lm.resids ~ nSpecies, data = plantbiomass, 
           ylab = "Residual",
           xlab = "Species number treatment",
           pch = 1,  # pch changes the symbol type
           col = "firebrick",
           las = 1)  # orients y-axis tick labels properly
abline(0, 0, lty = 2) # add horizontal dashed line 
}
```

Here we are watching out for:  

* a "funnel" shape, or substantial differences in the spread of Y at each X
* outliers to the general trend  
* a curved pattern  

> Figure 3 shows that the variance in biomass stability is greatest within the largest species richness treatment. This again suggests the need for a log transformation.

**TIP**:  If ever you see a clear outlier when checking assumptions, then a safe approach is to report the regression with and without the outlier point(s) included. If the exclusion of the outlier changes your conclusions, then you should make this clear.  

### Transform the data

Let's log-transform the response variable, creating a new variable `log.biomass`: 
```{r}
plantbiomass$log.biomass <- log(plantbiomass$biomassStability)
```

Now let's re-run the regression analysis and save the residuals again:
```{r}
log.biomass.lm <- lm(log.biomass ~ nSpecies, data = plantbiomass)
plantbiomass$log.biomass.lm.resids <- residuals(log.biomass.lm)
```

And re-plot the residual diagnostic plots:  

```{r fig.cap = "Figure 4: Normal quantile plot of the residuals from a regression of biomass stability (log transformed) on species number for 161 plots", fig.width = 4.2, fig.height=4}
{
qqnorm(plantbiomass$log.biomass.lm.resids, las = 1)
qqline(plantbiomass$log.biomass.lm.resids)
}
```

```{r fig.cap = "Figure 5: Residual plot from a regression of biomass stability (log transformed) on species number for 161 plots.", fig.width = 4.2, fig.height = 4}
{
plot(log.biomass.lm.resids ~ nSpecies, data = plantbiomass, 
           ylab = "Residual",
           xlab = "Species number treatment",
           pch = 1,  # pch changes the symbol type
           col = "firebrick",
           las = 1)  # orients y-axis tick labels properly
abline(0, 0, lty = 2) # add horizontal dashed line 
}
```

> Figure 4 shows that the residuals are reasonably normally distributed, and Figure 5 shows no strong pattern of changing variance in Y along X, and nor is there an obvious curved pattern to the residuals (which would indicate a non-linear association). We therefore proceed with the analyses using the log-transformed response variable.  

* * *

### Conduct the regression analysis  

We have already conducted the regression analysis above, using the `lm` function.  We stored the output in the `log.biomass.lm` object. Now, we use the `summary` function on the regression output to view the results:  

```{r}
summary(log.biomass.lm)
```

What do we need to focus on in this output?  

Under the "Estimate" heading, you see the estimated value of the Intercept (_a_) and of the slope (_b_), which is reported to the right of the predictor variable name.  

**NOTE**: There are two values of _t_ reported in the table; one associated with the intercept, and one with the slope. These are testing the implied hypotheses that (i) the intercept equals zero and (ii) the slope equals zero. We are typically only interested in the slope.    

At the bottom of the output you'll see a value of _F_, just like you saw with the ANOVA output in [Tutorial 15](https://people.ok.ubc.ca/jpither/modules/Comparing_more_than_2_means.html#conduct_the_anova_test). **It is this value of _F_ that you will report in statements about regression (see below).**  

We also see the _P_-value associated with the test of the zero slope null hypothesis, which is identical to the _P_-value reported for the overall regression (at the bottom of the output).  

**NOTE**: It is important to recognize that an overall significant regression (i.e. slope significantly different from zero) does not necessarily mean that the regression will yield accurate predictions. For this we need to assess the "coefficient of determination", or _R_^2^ value (called "multiple R-squared" in the output), which here is `r round(summary(log.biomass.lm)$r.squared, 2)`, and **represents the fraction of the variation in Y that is accounted for by the linear regression model**. A value of this magnitude is indicative of a comparatively weak regression, i.e. one that does has predictive power, but not particularly good predictive power (provided assumptions are met, and we're not extrapolating beyond the range of our X values).  

Before getting to our concluding statement, we need to additionally calculate the 95% confidence interval for the slope, $\beta$.  

* * *

### Confidence interval for the slope

To calculate the confidence interval for the true slope, $\beta$, we use the `confint` function from the base stats package:  

```
?confint
```

We run this function on the `lm` model output:  

```{r}
confint(log.biomass.lm)
```

This provides the lower and upper limits of the 95% confidence interval for the intercept (top row) and slope (bottom row).  

* * *

### Scatterplot with regression confidence bands

If your regression is significant, then it is recommended to accompany your concluding statement with a scatterplot that includes so-called **confidence bands** around the regression line.  

If your regression is not significant, **do not** include a regression line in your scatterplot!  

We can calculate two types of confidence intervals to show on the scatterplot:  

* the 95% "confidence bands"  
* the 95% "prediction intervals"  

Most of the time we're interested in showing **confidence bands**, which show the 95% confidence limits for the _mean_ value of Y in the population at each value of X. In other words, we're more interested in predicing an average value in the population (easier), rather than the value of an individual in the population (much more difficult).  

To display the confidence bands, we use the `visreg` function from the `visreg` package:  

```
?visreg
```

The function has a lot of options that we're going to ignore, becaue the default arguments are all good for our purposes.  

This "visreg" function takes the regression output object, here `log.biomass.lm`, and plots the regression line with confidence bands included. The argument "alpha" is used to set the level of confidence, and the default is alpha = 0.05, so 95% confidence bands. Note that you do not provide the original "X" and "Y" variables to the `visreg` function. Instead you provide the regression output object!  

Let's give it a try, making sure to spruce it up a bit with some additional options, and including a good figure caption, shown here for your reference (this is the code you'd put after the "r" in the header of the R chunk:  

```
fig.cap = "Figure 5. Stability of biomass production (log transformed) over 10 years in 161 plots and the initial number of plant species assigned to plots. Also shown is the significant least-square regression line (black solid line; see text for details) and the 95% confidence bands (grey shading)."
```

```{r fig.cap = "Figure 6: Stability of biomass production over 10 years in 161 plots and the initial number of plant species assigned to plots. Also shown is the significant least-square regression line (black solid line; see text for details) and the 95% confidence bands (grey shading).", fig.width = 4.2, fig.height = 4}
visreg(log.biomass.lm, alpha = 0.05, 
       line = list(col = "black"), 
       points = list(cex = 1.5, pch = 1,col = "black", lwd = 1.75), 
       las = 1, 
       xlab = "Species number treatment", 
       ylab = "Biomass stability (log transformed)")
```

Here we can interpret the figure as follows:  

> Figure 6 shows that biomass stability (log-transformed) is positively and linearly related to the initial number of plant species assigned to experimental plots, but there is considerable variation that remains unexplained by the least-square regression model.  

**NOTE**: Although the Guidelines for Data Presentation [document](https://people.ok.ubc.ca/jpither/more/UBCO_biology_guidelines_data_presentation_2016.pdf) indicates that you should report key information about the regression within the figure caption (e.g. the equation of the line), it is sufficient to refer the reader to the text.  

* * *

### Concluding statement

Now that we have our new figure including confidence bands (because our regression was significant), and all our regression output (including confidence intervals for the slope), we're ready to provide a concluding statement.  

**NOTE**: We simply report the sample size rather than the degrees of freedom in the parentheses.  

> As seen in Figure 6, species number is a significant predictor of biomass stability: _Biomass stability (log)_ = `r round(coef(log.biomass.lm)[1], 2)` + `r round(coef(log.biomass.lm)[2], 2)`(_Species number treatment_); _F_ = `r round(summary(log.biomass.lm)$fstatistic[1], 2)`; _n_ = `r length(na.omit(plantbiomass$biomassStability))`; 95% confidence limits for the slope: `r c(round(confint(log.biomass.lm)[2,1], 3), round(confint(log.biomass.lm)[2,2], 3))`; _R_^2^ = `r round(summary(log.biomass.lm)$r.squared, 2)`; _P_ < 0.001). Given the relatively low *R*^2^ value, predictions from our regression model will not be particularly accurate.   

**NOTE**: It is OK to report the concluding statement and associated regression output using the log-transformed data. However, if you wanted to make predictions using the regression model, it is often desirable to report the back-transformed predicted values (see below).   

* * *

## Model-I versus Model-II regression

Model-I regression, which we use here, assumes that there is negligible error or uncertainty associated with the "X" variable measurements, relative to that associated with the "Y" variable measurements. In other words, it effectively assumes that we have control over the "X" values (like we do in this biomass experiment). Model-II regression would be used in a situation where both X and Y have similar error or uncertainty, such as using measurements of height to predict values of weight among individuals.     

* * *

## Making predictions

Even though the *R*^2^ value from our regression was rather low, we'll proceed with using the model to make predictions.  

It is not advisable to make predictions beyond the range of X-values upon which the regression was built. So in our case (see Figure 6), we would not wish to make predictions using species richness values beyond 16.  Such "extrapolations" are inadvisable.  

To make a prediction using our regression model, use the `predict.lm` function from the base stats package:  

```
?predict.lm
```

If you do not provide new values of `nSpecies` (note that the variable name must be the same as was used when building the model), then it will simply use the old values that were supplied when building the regression model.  

Here, let's create a new set of `nSpecies` data to make predictions with.  For example, perhaps we wish to know what the predicted value of biomass stability would be if we were to grow plants in mixtures of 1 through 16 species, rather than just the five different values used in the experiment.  

A quirky thing with the `predict.lm` function is that we need to supply the exact same number of values of X that were used to build the model, and we also need to use the exact same variable name for X.  

So, let's use 161 values, all randomly selected from integers 1 through 16. Here's the code that creates a new data frame (`new.data`) and a variable `nSpecies` (it must be the same name as the original X variable) that holds our new values of "X" to use for predicting "Y".  

```{r}
set.seed(123)
new.X.data <- data.frame(nSpecies = sample(1:16, 161, replace = T)) # sampling with replacement 
```

Let's make the predictions, and we'll store the predicted values in our `plantbiomass` data frame, in a new variable:  

```{r}
plantbiomass$predicted.vals <- predict.lm(log.biomass.lm, newdata = new.X.data)
```

Let's superimpose these predicted values over the original scatterplot:

```{r fig.cap = "Figure 7: Stability of biomass production (log transformed) over 10 years in 161 plots and the initial number of plant species assigned to plots. Also shown are predicted values of Y across values of X.", fig.width = 4.2, fig.height = 4}
{
plot(log.biomass ~ nSpecies, data = plantbiomass, 
           ylab = "Biomass stability (log)",
           xlab = "Species number treatment",
           pch = 1,  # pch changes the symbol type
           col = "firebrick",
           las = 1)  # orients y-axis tick labels properly
points(new.X.data$nSpecies, plantbiomass$predicted.vals, pch = 16) # add predicted points 
}
```

Figure 7 shows pretty clearly the overal positive effect of species richness on biomass stability, and it also shows the considerable scatter in the relationship, hence the relatively low *R*^2^ value.  

### Back-transforming regression predictions 
 
Now remember that the predicted values above are **log-transformed** values, and it is often desirable to report predicted values back in their original non-transformed state. To do this, follow the instructions in the earlier [tutorial](https://people.ok.ubc.ca/jpither/modules/assumptions_transformations.html#back-transforming_log_data) dealing with data transformations. 

* * *

## List of functions

__Getting started__:

* `read.csv`
* `url`
* `library` 

__Data management / manipulation__: 

* `inspect` (`tigerstats` / `mosaic` packages)
* `head`

__Data visualization__:

* `plot`
* `jitter`

__Assumptions__:

* `qqnorm`
* `qqline`
* `residuals`
* `shapiro.test`
* `abline`

__Regression__:

* `lm`  
* `summary`  
* `confint`
* `visreg` (from the `visreg` package)

__Prediction__:  

* `predict.lm`
* `seq`
* `data.frame`
* `sample`
* `points`

* * *

## Markdown File

You can access the Rmd file that created this page [here](https://people.ok.ubc.ca/jpither/modules/least_squares_regression.Rmd). For it to work, you also need to download a "css" file linked [here](https://people.ok.ubc.ca/jpither/modules/tutorial.css) (called "tutorial.css"), and place it in a directory one down from your working directory.  




